---
title: "S&P 500 and Economic Indicators Analysis"
format: html
execute:
  echo: false
---

## Introduction
This project analyzes the relationship between the S&P 500 and key economic indicators. We utilized QuantMod as the package/library, to gather our data, to determine this correlation (causality, cointegration, and (PCA)), we utilized tests for causality, tests for correlation through PCA, and cointegration. Although some of these tests might not seem directly measure of correlation, they provide insight into the variables, that could be interpreted as a correlation

## Setup

Inside we have imported, various libraries, most important being quantmod, which contains our data.

```{r setup, message=FALSE, warning=FALSE, output=FALSE}
#| label: package-setup
#| include: false

# Install and load required packages
required_packages <- c("quantmod", "vars", "igraph", "ggplot2", "tidyverse", "roll", "lmtest", "urca")
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

suppressPackageStartupMessages({
  library(quantmod)
  library(vars)
  library(igraph)
  library(ggplot2)
  library(tidyverse)
  library(roll)
  library(lmtest)
  library(urca)
})
```
# Downloading Data
Here we download data from Yahoo Finance, (CPI), Federal Funds Rate (FFR)
```{r}
# Get S&P 500 data from Yahoo Finance
GSPC <- getSymbols("^GSPC", src = "yahoo", from = "1974-01-01", to = Sys.Date(), auto.assign = FALSE)
sp500_data <- data.frame(date = index(GSPC), coredata(GSPC))

# Download Consumer Price Index (CPI) data as a proxy for inflation
CPIAUCSL <- getSymbols("CPIAUCSL", src = "FRED", from = "1974-01-01", to = Sys.Date(), auto.assign = FALSE)
inflation_data <- data.frame(date = index(CPIAUCSL), coredata(CPIAUCSL))

# Download Federal Funds Rate data
FEDFUNDS <- getSymbols("FEDFUNDS", src = "FRED", from = "1974-01-01", to = Sys.Date(), auto.assign = FALSE)
interest_rate_data <- data.frame(date = index(FEDFUNDS), coredata(FEDFUNDS))
```

## Data Preperation
Inside of this section of the project, we merge the data into one data frame, check for missing data, and convert dates to Date type if needed.
```{r setup, warning=FALSE}
# Convert dates to Date type if necessary
sp500_data$date <- as.Date(sp500_data$date)
inflation_data$date <- as.Date(inflation_data$date)
interest_rate_data$date <- as.Date(interest_rate_data$date)

# Merge data into a single data frame
merged_data <- sp500_data %>%
  left_join(inflation_data, by = "date") %>%
  left_join(interest_rate_data, by = "date")

# Rename columns for clarity
colnames(merged_data) <- c("date", "SP500_Open", "SP500_High", "SP500_Low", "SP500_Close", "SP500_Volume", "SP500_Adjusted", "CPI", "Federal_Funds_Rate")

# Check for missing data and handle it
merged_data <- merged_data %>% 
  fill(CPI, .direction = "down") %>% 
  fill(Federal_Funds_Rate, .direction = "down")


```

# Granger Causality Analysis
```{r}
library(tseries)
library(vars)
library(lmtest)
library(tidyverse)

# Check for NA values

# Remove NA values
clean_data <- merged_data %>%
  select(date, Federal_Funds_Rate, SP500_Close) %>%
  na.omit()

# Verify data

# Check stationarity

# Calculate differences
clean_data$d_ffr <- c(NA, diff(clean_data$Federal_Funds_Rate))
clean_data$d_sp500 <- c(NA, diff(clean_data$SP500_Close))

# Remove the NA created by differencing
clean_data <- clean_data %>% na.omit()

# Granger causality with multiple lags
max_lag <- 12
granger_results <- data.frame(lag = 1:max_lag, p_value = NA)

for(lag in 1:max_lag) {
  granger_test <- grangertest(d_sp500 ~ d_ffr, order = lag, data = clean_data)
  granger_results$p_value[lag] <- tail(granger_test$`Pr(>F)`, 1)
}

# Plot results
ggplot(granger_results, aes(x = lag, y = p_value)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Granger Causality p-values: Federal Funds Rate → S&P 500",
       x = "Lag (months)", y = "p-value")

# Correlation analysis
cor_test <- cor.test(clean_data$d_ffr, clean_data$d_sp500)

# VAR model
var_data <- cbind(clean_data$d_ffr, clean_data$d_sp500)
var_model <- VAR(var_data, p = 1, type = "const")
```
This is an causality analysis employing a few tests, first is lags, due to the intervals between the relevant data, ie when the fed changes interest rates, or proceeds with a meeting. We can observe that the p-values consistently hover above 0.05 suggesting that there is no meaningful statistical evidence to establish      Granger causality, between the federal funds rate and the S&P. However, due to inherent limitations in computing, we only employed a linear analysis (basically just a fancy linear regression model) thus there could be a non-linear causality between the federal funds rate and the S&P 500.


# Principal Component Analysis

```{r}
pca_data <- merged_data %>% select(SP500_Close, CPI, Federal_Funds_Rate) %>% na.omit()
pca_result <- prcomp(pca_data, scale. = TRUE)

# Calculate loadings
loadings <- pca_result$rotation

# Create a biplot
biplot(pca_result, scale = 0)

```

This is a biplot from a Principal Component Analysis (PCA) of three variables: SP500_Close, CPI (Consumer Price Index), and Federal_Funds_Rate.

The axes represent the first two principal components (PC1 and PC2). The red arrows represent the original variables projected onto the PC space:

SP500_Close points towards the upper right CPI points towards the right Federal_Funds_Rate points towards the upper left

The length of the arrows indicates how well each variable is represented in this 2D projection. The angles between the arrows suggest correlations:

SP500_Close and CPI are positively correlated (acute angle) Federal_Funds_Rate is negatively correlated with both SP500_Close and CPI (obtuse angles)

The black dots represent individual data points projected onto this space, forming a U-shaped pattern.

This suggests that the S&P 500 closing price and CPI tend to move together, while the Federal Funds Rate tends to move in the opposite direction of both.

# Johansen Cointegration Test 
```{r}
# Prepare data for cointegration test
cointegration_data <- merged_data %>%
  select(SP500_Close, CPI, Federal_Funds_Rate) %>%
  na.omit()

# Perform Johansen Cointegration Test
johansen_test <- ca.jo(cointegration_data, type = "trace", K = 2, ecdet = "const", spec = "longrun")

# Extract test statistics and critical values
summary_jo <- summary(johansen_test)
test_stats <- summary_jo@teststat
crit_vals <- summary_jo@cval

# Create a data frame for plotting
cointegration_df <- data.frame(
  r = 0:2,
  test_statistic = test_stats,
  cv_10 = crit_vals[,1],
  cv_5 = crit_vals[,2],
  cv_1 = crit_vals[,3]
)

# Plot
ggplot(cointegration_df, aes(x = factor(r))) +
  geom_point(aes(y = test_statistic), color = "red", size = 3) +
  geom_line(aes(y = cv_10, group = 1), linetype = "dashed", color = "blue") +
  geom_line(aes(y = cv_5, group = 1), linetype = "dashed", color = "green") +
  geom_line(aes(y = cv_1, group = 1), linetype = "dashed", color = "purple") +
  labs(title = "Johansen Cointegration Test Results",
       x = "Number of Cointegrating Relations",
       y = "Test Statistic / Critical Values") +
  theme_minimal()

```
This graph shows the results of a Johansen Cointegration Test:

The x-axis represents the number of cointegrating relations (0, 1, or 2). The y-axis shows the test statistic and critical values. The red dots represent the test statistics for each number of cointegrating relations. The dashed lines represent critical values at different significance levels. We see that the test statistic (red dot) is above all critical value lines for 0 cointegrating relations, suggesting we reject the null hypothesis of no cointegration. For 1 cointegrating relation, the test statistic is below the critical values, suggesting we fail to reject the null hypothesis of at most 1 cointegrating relation.

This implies that there is likely one cointegrating relationship among the variables tested. This means that while the variables may individually be non-stationary, there exists a linear combination of them that is stationary, indicating a long-term equilibrium relationship between the variables.

## Conclusion
Our conclusion for our data is that the models predict that there could be a statistically significant correlation between the S&P 500 and CPI/FFR, however, it's more than likely that our models show a correlation that is non-linear, thus our tests understate the true correlation, as it discounts non-linear correlations.
